---
title: "Computing Sensitivity and Specificity Using collect_metrics"
author: 
  - "Francisco Cardozo"
  - "Catalina Cañizares"
date: 2023-07-22
categories: ["Sensitivity", "Tidymodels", "Specificity"]
tags: ["Sensitivity", "Tidymodels", "Specificity"]
---

(in progress)

In the last [post](https://focardozom.github.io/posts/2023-07-16-yarstick/2023-07-15.html), we discussed computing sensitivity and specificity using the `sensitivity()` and `specificity()` functions from the `yardstick` package. In this post, we'll demonstrate how to compute these metrics using the `collect_metrics` function from the `tidymodels` package.

### Dataset

First, let's recreate the confusion matrix from the previous post.

```{r recreate-confusion-matrix, warning=FALSE, message=FALSE}
library(tidyverse)  # Load tidyverse package
library(gt)         # Load gt package
library(gtsummary)  # Load gtsummary package

# Create a tibble with the confusion matrix data
confusion_matrix_long <- tibble(
  'Predicted Class' = c('1', '1', '0', '0'),
  'True Class' = c('True 1', 'True 0', 'True 1', 'True 0'),
  'Count' = c(40, 15, 10, 35)
)

confusion_matrix_long |> gt()
```

The `tbl_cross()` function is useful for computing these metrics after a bit of data wrangling.

```{r tbl_cross, warning=FALSE, message=FALSE}

# Duplicate each row of the tibble based on the value in the Count column
confusion_matrix_long |>
  uncount(Count) |>
  tbl_cross(percent = "col")  

```

The **sensitivity** of our model is 80%, and the **specificity** is 70% - pretty good values!

### Computing Sensitivity and Specificity Using collect_metrics

In machine learning, **sensitivity** and **specificity** are important metrics used to evaluate and improve model performance. When we tune model parameters across different datasets, it's useful to compute these metrics for each dataset. This gives us useful data to make decisions on enhancing model performance. We can calculate these metrics efficiently using the `collect_metrics()` function from the tidymodels package.

We'll use `fit_resamples()` to estimate the model many times. This uses different samples from our training data. Then, we'll use `collect_metrics()` to compute the **sensitivity** and **specificity** for each dataset.

### Cross-Validation Folds

We'll start by using `vfold_cv()` from tidymodels to create 10 folds of data. To get a better estimate of the sensitivity and specificity, we'll use the *repeats* argument to repeat the sampling process 10 times. This gives us 100 folds of data in total.

::: {.callout-tip}
## Cross-Validation 
You can learn more about cross-validation [here](https://www.tidymodels.org/start/resampling/).
:::

```{r vfold_cv, warning=FALSE, message=FALSE}
# Load packages
library(tidymodels) # Load tidymodels package
tidymodels::tidymodels_prefer() # Set tidymodels as the default modeling framework

set.seed(1906) # Set seed for reproducibility

# Create a tibble with the data
df <- confusion_matrix_long |>
  uncount(Count)  |> 
  rename(
    x = `Predicted Class`, 
    y = `True Class`
  )

# Create 10 folds of the data
folds <- vfold_cv(df, v = 10, repeats = 10)
```

### Model Specification

Having the folds ready, we can now specify the model. 

```{r fit_resamples, warning=FALSE, message=FALSE}

# Define a logistic regression model specification
model_spec <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

# Create a recipe for preprocessing the data
recipe_glm <- recipe(y ~ x, data = df) |>
  # Convert all nominal variables to dummy variables
  step_dummy(all_nominal(), -all_outcomes()) 

# Define a workflow for fitting the logistic regression model
wr_glm <- workflow() |>
  # Add the recipe to the workflow
  add_recipe(recipe_glm) |>
  # Add the model specification to the workflow
  add_model(model_spec)
```

Perfect! Now we are ready to estimate the model.

### Estimating the Model

```{r fit_resamples2, warning=FALSE, message=FALSE}

# Estimate the model 100 times
model_resamples <- fit_resamples(
  wr_glm, 
  folds, 
  control = control_resamples(save_pred = TRUE)
)
```

Notice that we used the *save_pred* argument in `control_resample()` because we want to save the predictions of the model for each fold. This allows us to compute the sensitivity and specificity for each fold, which is the goal of this post.

Let's take a look at one of the results.

```{r model_resamples, warning=FALSE, message=FALSE}

# Extract the metrics for the first resample
model_resamples |>
  # Select the first resample
  slice(1) |>
  # Unnest the metrics data
  unnest(.metrics)  |> 
  select(.metric, .estimate) |>
  gt()  |> 
  fmt_number(decimals = 3)
```

By default, we obtain **accuracy** and **roc_auc** metrics for classification models. However, we want a specific metric set (**sensitivy** and **specificity**). To achieve this, we can use `metric_set()` function from the `yardstick` package.

::: {.callout-caution}
## event_level = ?

Be mindful of the event level used by yardstick. As we demonstrated in the [previous post](https://focardozom.github.io/posts/2023-07-16-yarstick/2023-07-15.html), `yardstick` uses the first level as the event of interest. To change this behavior, we need to specify the *event_level* argument when using the `fit_resamples()` function.

:::


### Computing Sensitivity and Specificity

```{r warning=FALSE, message=FALSE}  

# Define the metrics to use
my_metrics <- metric_set(sens, spec)

# Fit the logistic regression model using resampling
model_resamples <- fit_resamples(
  wr_glm, 
  folds, 
  metrics = my_metrics,
  control = control_resamples(
    # Save the predicted values for each resample
    save_pred = TRUE,
    # Set the event level to "second"
    event_level = "second"
  )
)

```

::: {.callout-tip}
You can find all available metrics in yardstick [here](https://yardstick.tidymodels.org/reference/index.html).
:::

Now, let's dive into the results. For a start, we'll focus on the first resample. Although this may provide an imperfect estimation of our "true" values, theory assures us that as we increase the number of resamples, the average of the metrics will move closer to these "true" values. Let's see this in action!

```{r}  
# Extract the metrics for the first resample
model_resamples %>%
  # Select the first resample
  slice(1) %>%
  # Unnest the metrics data
  unnest(.metrics) %>%
  # Select the metric and estimate columns
  select(.metric, .estimate) %>%
  # Create a gt table
  gt() %>%
  # Format the estimate column to 3 decimal places
  fmt_number(decimals = 3)

```

Good, we have achieved a **sensitivity** of 100% and **specificity** of 57.1%, which, frankly, isn't an ideal representation of our true values. But, let's keep in mind, this is just a single estimate out of our 100 trials. Now, let's take it to the next level and summarize these values across all our estimates.

```{r}  
# Collect and summarise the metrics for all resamples
model_resamples %>%
  # Collect the metrics and summarise them
  collect_metrics(summarise = TRUE) %>%
  # Create a gt table
  gt() %>%
  # Format the table to 3 decimal places
  fmt_number(decimals = 3)
```

Great job! Our model's estimated **sensitivity** is at 81.0% and **specificity** stands at 70.3%, quite near to our actual values.

With all this valuable data at hand, it's tempting to visualize it. So, let's move on to plotting the observed distribution for both sensitivity and specificity values to get a better understanding of our model's performance. 

```{r warning=FALSE, message=FALSE}  

model_resamples  |> 
  unnest(.metrics)  |> 
  ggplot(aes(x = .estimate, fill = .metric)) +
  geom_density() +
  facet_wrap(~.metric) +
  theme_minimal() +
  scale_fill_manual(values = c("#00AFBB", "#E7B800"))
   
```

Next time, we'll look more closely at these values. We'll work out which ones are bigger than 0.5 and see how spread out they are. But, we need to be careful because these numbers are connected; they all come from the same data. Don't miss it!

## Summary 

In this blog post, we discussed the use of the `tidymodels` package to compute the sensitivity and specificity of a model - two important measures for evaluating model performance. We started with the recreation of a confusion matrix, then used the `tbl_cross()` function to perform some data wrangling.

Next, we leveraged `fit_resamples()` from the `tidymodels` package to estimate our model multiple times, using cross-validation with 10 repeated folds to create 100 datasets in total.

We set up a logistic regression model, fit it to our data, and estimated the model multiple times using the created folds. The saved predictions allowed us to compute sensitivity and specificity for each fold.

We defined a custom set of metrics - roc_auc, sensitivity, and specificity - using the `metric_set()` function from the `yardstick` package. Finally, we used `collect_metrics()` to compute the average of these metrics for all the folds, and visualized the distribution of the sensitivity and specificity values.

Future posts will delve into calculating confidence intervals and evaluating the variability of these values across resamples.
