---
title: "How to Create a Bash Script to Extract Bibliographic References from DOIs"
author: Francisco Cardozo
format: html
date: 2024-04-06
categories: ["DOI", "Bash", "References"]
tags: ["DOI", "Bash", "References"]
---

Dealing with references can feel like a real headache. However, recently I came across a fantastic solution by [Richard Mclearn](https://twitter.com/rlmcelreath). This blog post is all about making your life easier with a simple tutorial for Mac users. I'll show you how to set up a `getref` function that quickly grabs bibliographic information in BibTeX format from a DOI or a DOI URL.

![[source](https://twitter.com/rlmcelreath/status/1767812801404223535)](x-image.png){width=70%}

# Step-by-Step Guide

## Step 1: Prepare Your Workspace

First, we need to create a dedicated folder to store your `getref` function. This not only helps in organizing your scripts but also in managing your PATH environment efficiently.

You can create the folder following this steps:

1.1. Open the terminal (cmd+space then type terminal).

1.2. Create a new folder named bin in your home directory: 

type 

```{r eval=FALSE}  
mkdir ~/bin
```

and press return in your keyboard.

:::{.callout-note}
I created the bin folder in my home directory. It will make it easier to access the script later on.
:::

## Step 2: Move to the bin Folder

Navigate to the newly created folder by typing 
```{r eval=FALSE}  
cd ~/bin
```

## Step 3: Create the Script File

Create a new file named getref by typing 
```{r eval=FALSE}  
touch getref
```

This file will be our script.

## Step 4: Add the Script Code

Open the `getref` file in your favorite text editor. For simplicity, you can use  [Nano](https://www.nano-editor.org).

```{r eval=FALSE}  
nano getref
```

In the inside of the file, copy and paste the following code snippet. This script checks if a DOI or DOI URL is provided as an argument and then fetches the bibliographic information in BibTeX format.

```{r eval=FALSE}
#!/bin/bash
# Check if an argument is provided
if [ "$#" -ne 1 ]; then
    echo "Usage: $0 <DOI or DOI URL>"
    exit 1
fi

# Determine if the input is a DOI URL or just a DOI number
if [[ "$1" =~ ^https:// ]]; then
    # Extract the DOI number from the URL
    DOI=$(echo "$1" | sed 's|https://doi.org/||')
else
    # Assume the input is just a DOI number
    DOI="$1"
fi

# Extract and print the reference
curl -LH "Accept: application/x-bibtex" "https://doi.org/$DOI"
```

## Step 5: Make the Script Executable

Ensure the script can be executed by running in the terminal

```{r eval=FALSE}  
chmod +x ~/bin/getref
```

:::{.callout-note}
Notice that I used the path to my file, which is located in the bin folder in my home directory.
:::

## Step 6: Update Your Shell Configuration

Now, Open your shell configuration file (.zshrc for Zsh or .bashrc for Bash) using a text editor, e.g., 

```{r eval=FALSE}  
nano ~/.zshrc
```

You will need to add **~/bin** to your **PATH**. To do this, add the following line at the end of the file:

```{r eval=FALSE}
export PATH="$HOME/bin:$PATH"
```

Save and close the file.

## Step 7: Apply the Changes

Apply the changes to your current session by sourcing your configuration file: 

```{r eval=FALSE}  
source ~/.zshrc
```

for Zsh 

or 

```{r eval=FALSE}  
source ~/.bashrc
```

for Bash.

# How to Use Your New Script

To use your script, simply type `getref` in your termina followed by a DOI or a DOI URL. For example:

```{r eval=FALSE}
getref 10.1080/09687637.2023.2167649
```

or

```{r eval=FALSE}
getref https://doi.org/10.1080/09687637.2023.2167649
```

The script will fetch and display the bibliographic information in BibTeX format.

**Done!** This is a handy tool that simplifies the process of fetching bibliographic data. I hope you enjoy it. If you have any questions or suggestions, feel free to leave a comment below.

